---
title: "1.3.2: Data Wrangling"
subtitle: "(Asynchronous-Online)"
bibliography: ./packages.bib
nocite: |
  @*
format:
  html: default
  pdf: default
editor_options: 
  chunk_output_type: console
---

\thispagestyle{fancy}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Session Objectives

1. To read in data.
2. To subset data.
3. To select, create, and modify variables.
4. To filter the observations according to a certain condition.
5. To explore and summarize data.
6. To run descriptive statistics.

Key points to cover:
1.	Import data.
2.	Introduce to tidyverse.
3.	Summarize data and run descriptive statistics.
4.	Introduce other resources (e.g., books, blogs, or websites) trainees can refer to.

---

## 0. Prework - Before You Begin

### Install Packages

Before you begin, please go ahead and install the following packages - these are all on CRAN, so you can install them using the RStudio Menu Tools/Install Packages interface:

* haven
* readr
* readxl
* dplyr







---

\newpage

## 1. To read in data.

### Begin with a NEW Project

add steps and screenshot on starting with a new project...

### importing data


### exploring builtin datasets - see environment also - including data with packages

---

\newpage

## 2. To subset data.



---

\newpage

## 3. To select, create, and modify variables.





---

\newpage

## 4. To filter the observations according to a certain condition.



---

\newpage

## 5. To explore and summarize data.

### base r - summary stats

### making tables

### gtsummary package

### other packages - arsenal, tableone and Rmarkdown formatting of tables... get inspired

table competitions

---

\newpage

## 6. To run descriptive statistics.






---

\newpage

```{r echo=FALSE}
knitr::write_bib(x = c(.packages()), 
                 file = "packages.bib")
```

## References

::: {#refs}
:::

## Other Helpful Resources

[**Other Helpful Resources**](./additionalResources.html)

