---
title: "1.3.1: Introduction to the R and R Studio (Asynchromous-Online)"
format:
  html: default
  pdf: default
editor_options: 
  chunk_output_type: console
---

## Session Objectives

1. Get aquainted with R and R Studio
2. Write simple R code in Console
3. Create your first R script 
4. Install and load R packages (understand R session)
5. Create your first R Markdown report and produce output files in different formats (HTML, PDF, or DOCX)

## Prework - Before You Begin

some info

::: callout-note
R is the name of the programming language itself and RStudio is a convenient interface, commonly referred to as an integrated development environment or an IDE, for short.
:::

## 1. Get aquainted with R and R Studio

info here 

## 2. Write simple R code in Console

info here 

## 3. Create your first R script 

info here 

## 4. Install and load R packages (understand R session)

info here 

## 5. Create your first R Markdown report and produce output files in different formats (HTML, PDF, or DOCX)


some info

### section 1

summary function

```{r}
summary(cars)
```

### figures

making simple plots

::: {.panel-tabset}

## Code

```{r eval=FALSE}
library(ggplot2)
library(dplyr)

ggplot(pressure, aes(temperature, pressure)) +
  geom_point() +
  geom_line(color = "blue") +
  ggtitle("My plot title")
```

## Plot

```{r echo=FALSE}
library(ggplot2)
library(dplyr)

ggplot(pressure, aes(temperature, pressure)) +
  geom_point() +
  geom_line(color = "blue") +
  ggtitle("My plot title")
```

:::

