---
title: "1.3.5: Statistical Tests and Models"
subtitle: "(In Person)"
bibliography: ./packages.bib
nocite: |
  @*
format:
  html: default
  pdf: default
editor_options: 
  chunk_output_type: console
---

\thispagestyle{fancy}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      message = FALSE,
                      warning = FALSE)
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
# be sure to cover these
# https://easystats.github.io/easystats/
# and tidymodels, broom, purrr, etc
# effect sizes
```

## Session Objectives

1. Develop linear and logistic regression models.
2. (Use a survey sampling weight to generate more representative descriptive and inferential statistical values.) - Currently, this objective is under the Module 1.3.4: Missing data and sampling weight.
3. Interpret a model output.

---

## 0. Prework - Before You Begin

### A. Install packages

If you do not have them already, install the following packages from CRAN (using the RStudio Menu "Tools/Install" Packages interface):

* [`VIM`](https://cran.r-project.org/web/packages/VIM/index.html){target="_blank"} and [`VIM` package website](https://statistikat.github.io/VIM/index.html){target="_blank"}
* (Optional) [`skimr`](https://cran.r-project.org/web/packages/skimr/index.html){target="_blank"} and [`skimr` website](https://docs.ropensci.org/skimr/){target="_blank"}
* (Optional) [`modelsummary`](https://cran.r-project.org/web/packages/modelsummary/){target="_blank"} and [`modelsummary` website](https://modelsummary.com/){target="_blank"}
* (Optional) [`summarytools`](https://cran.r-project.org/web/packages/summarytools/){target="_blank"} and [`summarytools` on Github](https://github.com/dcomtois/summarytools){target="_blank"}
* [`palmerpenguins`](https://cran.r-project.org/web/packages/palmerpenguins/){target="_blank"} and [`palmerpenguins` website](https://allisonhorst.github.io/palmerpenguins/){target="_blank"}
* [`ggplot2`](https://cloud.r-project.org/web/packages/ggplot2/){target="_blank"} and [`ggplot2` website](https://ggplot2.tidyverse.org/){target="_blank"}
* [`naniar`](https://cran.r-project.org/web/packages/naniar/){target="_blank"} and [`naniar` website](https://naniar.njtierney.com/){target="_blank"}
* [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) and [`dplyr` website](https://dplyr.tidyverse.org/)
* [`gtsummary`](https://cran.r-project.org/web/packages/gtsummary/) and [`gtsummary` website](https://www.danieldsjoberg.com/gtsummary/)
* [`Hmisc`](https://cran.r-project.org/web/packages/Hmisc/) and [`Hmisc` website](https://hbiostat.org/r/hmisc/)
* [`mice`](https://cran.r-project.org/web/packages/mice/index.html){target="_blank"} and [mice website](https://amices.org/mice/){target="_blank"}

### B. Review these online Book Chapters:

* [BOOK: Flexible Imputation of Missing Data, 2nd ed., by Stef van Buuren (`mice` package author) - Chapter 1 "Introduction", Sections 1.1-1.4](https://stefvanbuuren.name/fimd/ch-introduction.html){target="_blank"}
* [BOOK: The Epidemiologist R Handbook - Chapter 20 "Missing Data"](https://www.epirhandbook.com/en/new_pages/missing_data.html)

### C. Open/create an RStudio project for this lesson

Let's start with the `myfirstRproject` RStudio project you created in [Module 1.3.2 - part 1](module132_DataWrangling.html#begin-with-a-new-rstudio-project). If you have not yet created this `myfirstRproject` RStudio project, go ahead and create a new RStudio Project for this lesson. _Feel free to name your project whatever you want, it does not need to be named `myfirstRproject`._

---

\newpage

## 1. Develop linear and logistic regression models.

### Linear Regression

One aaaaaaaaaaaaaaa

### Logistic Regression

xxxxxxxxxxxxxxxxxx

---

\newpage

## 2. Linear and Logistic Regression Models with complex survey sampling weights

aaaaaaaaaaaaa

---

\newpage

## 3. Interpret a model output

aaaaaaaaaaaaa

---

\newpage

```{r echo=FALSE}
knitr::write_bib(x = c(.packages()), 
                 file = "packages.bib")
```

## R Code For This Module

* [`module_135.R`](module_135.R)

## References

::: {#refs}
:::

## Other Helpful Resources

**Missing Data Resources**

* [CRAN Task View for Missing Data](https://cran.r-project.org/web/views/MissingData.html)
* [R-miss-tastic Website](https://rmisstastic.netlify.app/)
* [Flexible Imputation of Missing Data (online book for 2nd edition) by Stef van Buuren](https://stefvanbuuren.name/fimd/)
* [Blog post on Missing Data Visualization in R using ggplot2](https://www.datawim.com/post/missing-data-visualization-in-r/)
* [Missing data R tutorial](https://libguides.princeton.edu/R-Missingdata)
* [CRAN Task View on Missing Data](https://cran.r-project.org/web/views/MissingData.html)
* [A resource website on missing values](https://rmisstastic.netlify.app/)
* [Handling missing values with R - tutorial](https://rmisstastic.netlify.app/tutorials/josse_tierney_bookdown_user2018tutorial_2018)
* [Blog post "My favourite R package for: summarising data"](https://dabblingwithdata.amedcalf.com/2018/01/02/my-favourite-r-package-for-summarising-data/)

and

[**Other Helpful Resources**](./additionalResources.html){target="_blank"}







