---
title: "1.3.3: Data Visualization"
subtitle: "(Asynchronous-Online)"
bibliography: ./packages.bib
nocite: |
  @*
format:
  html: default
  pdf: default
editor_options: 
  chunk_output_type: console
---

\thispagestyle{fancy}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Session Objectives

1. To visualize data using different R packages.



Key points to cover:
1.	Introduce to ggplot2 and other R packages.
2.	Visualize one, two, or more variables at a time.
3.	Introduce other resources (e.g., books, blogs, or websites) trainees can refer to.



---

## 0. Prework - Before You Begin



---

\newpage

## 1. Base R graphical functions


---

\newpage

## 2. ggplot2 package



---

\newpage

## 3. Get boilerplate code to start

### R Gallery


### R Graphics Cookbook






---

\newpage

```{r echo=FALSE}
knitr::write_bib(x = c(.packages()), 
                 file = "packages.bib")
```

## References

::: {#refs}
:::

## Other Helpful Resources

[**Other Helpful Resources**](./additionalResources.html)

