---
title: "1.3.6: Putting Reproducible Research Principles Into Practice"
subtitle: "(Asynchronous-Online)"
bibliography: ./packages.bib
nocite: |
  @*
format:
  html: default
  pdf: default
editor_options: 
  chunk_output_type: console
---

\thispagestyle{fancy}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Session Objectives

1. Discuss reproducible research principles.
2. Apply reproducible research principles to data analysis using R Markdown.


Key points to cover:
1.	Reproducible research principles
2.	What is R Markdown
3.	How to create a report using R Markdown
a.	Customize the layout of presentations or reports
b.	Insert and create objects, such as tables, images, or videos, within a document






---

## 0. Prework - Before You Begin






---

\newpage

## 1. Discuss reproducible research principles.



---

\newpage

## 2. Apply reproducible research principles to data analysis using R Markdown.





---

\newpage

```{r echo=FALSE}
knitr::write_bib(x = c(.packages()), 
                 file = "packages.bib")
```

## References

::: {#refs}
:::

## Other Helpful Resources

[**Other Helpful Resources**](./additionalResources.html)

