---
title: "1.3.6: Putting Reproducible Research Principles Into Practice"
subtitle: "(In Person)"
bibliography: ./packages.bib
nocite: |
  @*
format:
  html: default
  pdf: default
editor_options: 
  chunk_output_type: console
---

\thispagestyle{fancy}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      message = FALSE,
                      warning = FALSE)
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6
)
```

## Session Objectives

1. Discuss reproducible research principles.
2. Apply reproducible research principles to data analysis using R Markdown.

Key points that will be covered:

1. Reproducible research principles
2. What is R Markdown
3. How to create a report using R Markdown
    - Customize the layout of presentations or reports
    - Insert and create objects, such as tables, images, or videos, within a document

### In Person - Summer 2025

During the in person session we will work through creating and using Rmarkdown to combine report text, R code and output seamlessly to create reproducible analysis reports.

We will be working with the [PRAMS Report example](prams_report.html) in class. Also see complete details in the [PRAMS Module - section 7 on Reproducible Research Report](prams#prams-reproducible-research-report).

---

\newpage

```{r echo=FALSE}
knitr::write_bib(x = c(.packages()), 
                 file = "packages.bib")
```

## RMarkdown Code For This Module

* [`prams_report.Rmd`](prams_report.Rmd)

## References

::: {#refs}
:::

## Other Helpful Resources

[**Other Helpful Resources**](./additionalResources.html){target="_blank"}


